---
title: "Using Particle Swarm Optimization Algorithms for Finding E-optimal Design of Heteroscedastic Polynomial Model"
author: "Ping-Yang Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# It is a Draft!
# Theoretical Foundation

We revisit @dette1993, where he found selected $E$-optimal designs for polynomial models with response $y$ and its expectation, $E(y)$, is modeled as a polynomial regression  model using a single explanatory factor with $M$ monomials, $p_1, p_2, \ldots, p_M$, i.e.
\begin{equation}
E(y) = \eta(x, \boldsymbol\theta) = \theta_0 + \sum_{m = 1}^M \theta_m x^{p_m},~x\in[a, b],~\boldsymbol\theta = (\theta_0, \theta_1, \ldots, \theta_M).
\end{equation}
The simplest case is obtained when $(p_1, p_2, \ldots, p_M) = (1, 2, \ldots, M)$, and the regression function becomes a polynomial of degree $M$.

\cite{dette1993} considered polynomial models with heteroscedastic errors where the variance of the response at $x$ is modeled
as  $\sigma^2/\lambda(x)$, where $\lambda(x)$ is a user-specified efficiency function defined on the design space, $[a, b]$.  He focused on efficiency function of the form 
\begin{equation}
\lambda(x) = (x - a)^u (b - x)^v,
\end{equation}
where $u$ and $v$ are known constants.

# Implementation


## Dependency

```{r pkgins, eval = FALSE}
install.packages("globpso")
install.packages("Rcpp")
install.packages("RcppArmadillo")
```

## `genHeterPolyE`: The Main Function

| Input Parameters | Illustration                                                                                            | Default Value |
| ---------------- | ------------------------------------------------------------------------------------------------------- | ------------- |
| `nSupp`       | A integer number of the number of support points satisfying `nSupp >= 2`.                                       | `2`        |
| `designSpace` | The vector of the design space, `c(lower, upper)`, satisfying `lower < upper`.                                  | `c(-1, 1)` |
| `polyOrder`   | The vector of the order of polynomial model terms.  The default is \code{c(0, 1)}.                              | `c(0, 1)`  |
|               | Set `0` for intercept term and `-Inf` for $\log(x)$ term. Otherwise, set `a` as numerical value for $x^a$ term. |            |
| `varParam`    | The vector of the variance parameters, \code{c(u, v)} and \code{u >= 0} and \code{v >= 0}.                      | `c(1, 1)`  |
| `designType`  | The character of type of optimal design criterion including \code{"D"}, \code{"A"} or \code{"E"}.               | `"E"`      |
| `eqvData`     | The logical value controls if the function returns the data for drawing the equivalence plot.                   | `200`      |
| `nswarm`      | A integer number of swarm size in PSO algorithm.                                                                | `32`       |
| `maxiter`     | A integer number of maximal PSO iterations.                                                                     | `100`      |
| `seed`        | The random seed that controls initial swarm of PSO.                                                             | `NULL`     |
| `verbose`     | The logical value controls if PSO would reports the updating progress.                                          | `TRUE`     |


# Examples

```{r import}
# Import the main function
source("Approx_HeterPoly.R")
```

## Case 1

```{r case1set}
# Set the design requirements
n <- 2                # the number of support points
dSpace <- c(-1, 1)    # the design space, c(lower, upper)
pOrder <- c(0, 1)     # the order of polynomial model terms
                      # 0: intercept
                      # -Inf: log(x)
                      # a: x^a, a is numerical value
vParam <- c(1.5, 0.5) # the variance parameters, c(u, v)
```

```{r case1run}
# Run PSO for Finding E-optimal Design of Heteroscedastic Polynomial Model
out <- genHeterPolyE(nSupp = n, designSpace = dSpace, polyOrder = pOrder, 
              varParam = vParam, designType = "E", 
              eqvData = TRUE, nswarm = 32, maxiter = 100, seed = NULL, verbose = FALSE)
```

```{r case1res}
# The PSO-generated Design
out$optDesign

# The corresponding E-criterion value 
# (NEGATIVE of the minimal eignevalue of FIM)
out$CritVal

# Computing time
out$pso_results$cputime
```

```{r case1plot}
# Draw the equivalence plot
plot(out$Equiv$grid, out$Equiv$DD, type = "l", col = "blue")
abline(h = 0, lty = 3)
```

## Case 2

Statistical packages like STATA and SAS also provide fitting routines using fractional polynomials, suggesting that fractional polynomials are in demand in practice.  
Consider, for example, we are interested to find a $E$-optimal design for a fractional polynomial regression model on the interval $[20,80]$, where the variance of the univariate response $y$ is inversely proportional to a known function of the form $\lambda(x)=(x-20)^u(80-x)^v$, where $u, v$ are supposed to be known and different choices of $u$ and $v$ shapes the variance of the response across the design interval. The model is
\begin{equation}
E(y) = \theta_0 + \theta_1x^{-2} + \theta_2x^{-1} + \theta_3x^{-0.5} + \theta_4\log{(x)}
\end{equation}

```{r case2set}
# Set the design requirements
n <- 5                # the number of support points
dSpace <- c(20, 80)    # the design space, c(lower, upper)
pOrder <- c(0, -2, -1, -0.5, -Inf)     # the order of polynomial model terms
# 0: intercept
# Inf: log(x)
# a: x^a, a is numerical value
vParam <- c(1.5, 0.5) # the variance parameters, c(u, v)
```

```{r case2run}
# Run PSO for Finding E-optimal Design of Heteroscedastic Polynomial Model
out <- genHeterPolyE(nSupp = n, designSpace = dSpace, polyOrder = pOrder, 
                     varParam = vParam, designType = "E", 
                     eqvData = TRUE, nswarm = 128, maxiter = 100, seed = NULL, verbose = FALSE)
```

```{r case2res}
# The PSO-generated Design
out$optDesign

# The corresponding E-criterion value 
# (NEGATIVE of the minimal eignevalue of FIM)
out$CritVal

# Computing time
out$pso_results$cputime
```

```{r case2plot}
# Draw the equivalence plot
plot(out$Equiv$grid, out$Equiv$DD, type = "l", col = "blue")
abline(h = 0, lty = 3)
```

# Reference
